Fused regression for multi-source gene regulatory network inference

### 2016-01-17 paper1.tex
zackedits

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become attainable \cite{bonneau_predictive_2007, ciofani_validated_2012, carro_transcriptional_2010}. Large-scale data collection efforts have contributed to the development of high quality networks which accurately recapitulate biological processes, but there remain many processes and organisms for which there does not exist adequate data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine data from multiple platforms, lest we lose valuable information from existing studies. The problem of inferring related -- but not necessarily identical -- structure from related -- but not identical -- data is ubiquitous in biology. Multi-source network inference has applications for learning multiple networks in multiple related species, for learning networks associated with distinct processes within the same species, and for learning networks based on heterogenous data sources.

Moreover, as it becomes possible to learn genome-wide regulatory networks, we can begin to compare and to test whether there is conservation of networks across species and biological processes. Our use of model organisms to study biological processes and diseases relevant to humans relies on the assumption of conservation; yet this has not been tested at the genome scale.

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference \cite{bonneau_predictive_2007, arrieta-ortiz_experimentally_2015}. The core of both methods is the observation that biological information about the relatedness of genes can be used to constrain which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our approach uses predictions of similarity of putative interactions, to improve our ability to accurately describe TF-gene relationships on a genome-wide scale.

Our first method -- fused ridge -- uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when the overlap between genes is incomplete. Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously infer the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different). With this approach, we seek to test the hypothesis of conservation of interactions across networks.

@@

Existing multi species approaches often use orthology as a proxy for functional conservation \cite{penfold_inferring_2015, joshi_multi-species_2015, kashima_simultaneous_2009, zhang2010nearly}, or attempt to learn functional similarity via expression data \cite{gholami_cross-species_2010}. Orthology -- the measure of gene similarity we use -- can be approximated using readily identifiable sequence similarity. Sequence similarity is often a useful predictor of functional similarity \cite{wilson_assessing_2000}. In multi-species network inference, our fused L2 approach minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with evolutionary constraints created using orthology. However, we recognize that some genes may have taken on different functions and therefore may have new regulatory interactions. Gene duplications and neofunctionalization in multi-species network inference \cite{eisen_phylogenomics:_1998}, and changes in chromatin configuration in multi-cell line network inference \cite{li_role_2007}, are cases where assumptions about functional conservation may not hold.

@@

We approach this problem by introducing a saturating penalty function based on work in statistics to develop unbiased regularization penalties for regression \cite{zhang2010nearly, fan2001variable}. The main difference between the L2 fusion approach and this adaptive fusion approach occurs when the difference between presumed analogous interactions is large despite the fusion penalty: we assume that large differences represent evidence that gene functions have diverged and represent this, in adaptive fusion, with a relaxation of the fusion penalty. The resulting cost function is non-convex and difficult to optimize; however, we can approximate its solution and obtain deeper insight into functional similarity than is available through orthology or the comparison of separately learned networks.

@@

Information about the conserved structure of gene regulation is introduced through the incorporation of constraints into the above regression formulation. These constraints penalize differences between interaction weights in the networks of multiple species that are expected to be similar based on prior biological knowledge. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $p_{\lambda_S}(\beta^A_{g,k} - \beta^B_{h,l})$ in the quantity being minimized in order to encourage similarity. The function $p_{\lambda_S}(x)$ controls the shape of the relationship between weight dissimilarity and penalty, while parameter $\lambda_S$ controls the overall scaling of the penalization of differences between fused coefficients. $\lambda_S$ controls the tradeoff between fitting the expression data alone and producing a set of networks that conform to evolutionary prior knowledge. This gives us the final equation to be minimized 
\begin{equation}
%\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} p_{\lambda_S}(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})
\argmin_{\beta^S} \displaystyle\sum_{S \in \text{species}} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\substack{(g,h) \in orth,\\
 (h,l) \in orth}}x p_{\lambda_S}(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})
\end{equation}

where the second sum is over pairs of interactions with fusion constraints. In the fused L2 algorithm we develop, the penalty function is equal to the L2 norm of the difference in regulatory weight of fused coefficients. That is, $p_{\lambda_S}(x)=x^2$. The overall objective function for network inference is then:

@@

Because every component of the objective function is an L2 norm, the entire problem is convex, and can be solved through linear regression with a suitable design matrix. %The regulators of a pair of genes can be solved independently as long as the pair is not linked by fusion constraints, or a chain of fusion constraints; thus the problem of cross-species GRN inference decomposes into a large number of manageable regression problems as long as the orthology mappings are sparse.

In order to better understand the behavior of this objective function, consider the case where there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism). The choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). Part of the appeal of the approach, however, is that it allows pooling of data even when there is incomplete orthology. By introducing constraints on the similarity of individual interactions, rather than on the networks as a whole \cite{liu2011temporal}, we can pool some information across species even when an arbitrarily small fraction of genes have orthologs. This is particularly useful when dealing with a large number of species; pairwise orthologies may be nearly complete even when the number of genes present in every organism is small.

\subsection{Statistical approach and background}
Most network inference problems have more predictors than experimental samples ($p >> N$). Many methods have been proposed for penalized regression, such as ridge regression \cite{hoerl_ridge_1970}, and applied to genomic data \cite{waldron_optimized_2011, li_network-constrained_2008}. In our application, we consider the case where related outputs (genes) are likely to be affected by common inputs (transcription factors). Fusion has been proposed as a method for utilizing correlation information from data by penalizing pairwise differences of coefficients \cite{Daye2009,Price2014}. Approaches using fusion combine statistical strength across the multiple regression tasks by introducing regularization functions reflecting prior knowledge on output structure representing how outputs are related \cite{kim_tree-guided_2012,Land1997,Chen2010,Petry2011,Hebiri2011}. Price et al. and Bilgrau et al. use a fused ridge estimator for jointly estimating multiple inverse covariance matrices \cite{Price2014a,Bilgrau2015}; we take a similar approach, adding an fused L2 penalty to our ridge regression to incorporate prior knowledge about relationships between input-output pairs. Sohn et al. attempt to simultaneously learn the regression coefficients and the output structure \cite{sohn_joint_2012}. We also take a simultaneous approach, where we apply a penalty function bounded by a constant to produce unbiased estimators for large coefficients, combined with an L2 penalty, similar to SCAD-L2 \cite{Zeng2012}. This approach allows us to simultaneously learn the regression coefficients and evaluate the validity of our prior information.

@@

We use the fact that ridge constraints can be incorporated into a least-squares regression problem by appending a scaled identity matrix to the design matrix, and a corresponding number of zeros to the response vector. Similarly, a fusion constraint $\lambda_S (\beta_{i} - \beta{j})^2$ can be incorporated into a least-squares regression problem by appending a row containing $\sqrt{lambda_S}$ in the $i$th position, $-\sqrt{lambda_S}$ in the $j$th position, and $0$s elsewhere to the design matrix, and zero to the response vector.

In order to convert an optimization over multiple response variables (indexed by $i$) and multiple sources (indexed by $j$) with design matrices $X^{S_j}$ and response variables $y_i^{S_j}$ into an optimization with a single source and response variable, we vectorize as follows. Let $G$ be a set containing $(i,j)$ for each of the data sources and response variables currently being optimized. We construct the matrix $A$ by diagonally concatenating each of $X^{S_j}$ (with the remaining entries 0) and vertically concatenating each of $y_j^{S_j}$. We then solve
%Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
%$y =$ vertically concatenate the columns of $Y$. We solve

@@

%We can then define $W_{i,j} = \beta_{(j-1) \times N + i, (j-1) \times M + j}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication. In an ordinary regression problem each response variable can be solved independently as a large number of small regression problems. However, in fused regression, we append additional rows to the design matrix $A$ that link entries of $W$ associated with different response variables. As a result, these linked response variables must be solved simultaneously through vectorization. Two response variables are linked by a fusion constraint if any of the regulatory weights affecting those genes are linked by a fusion constraint. Two response variables must be solved simultaneously if there is any chain of linked response variables connecting them. However, every other response variable can be solved separately. In biological terms, the regulators of two genes (whether in the same species, or different species) must be solved together if there is a fusion constraint linking those genes' regulators, or if there is a chain of such constraints. This is important because vectorization produces a design matrix with a number of rows and columns equal to the sum of the number of rows and columns of the design matrices associated with each of the response variables (genes) being solved for. If the networks for a large number of genes are solved simultaneously, the system of equations can quickly become intractible to solve.

In order to avoid this difficulty, we use depth-first search to identify linked columns of each TF expression matrix $Y^{S_j}$, then forming design and response matrices through vectorization. We can then incorporate fusion constraints as in the case of single-source single response-variable fused regression. In most cases, we have found the direct solve using augmented matrices to be adequate. This is because the structure of orthology tends to be sparse, so only a small number of genes must be solved at once. However, in the general case of $c_i$ response variables from each of $i \in \text{species}$ data sources the design matrix to be solved is of size $\displaystyle \sum_{c_i} rows(X^{S_i}) \times \displaystyle \sum_{c_i} cols(X^{S_i})$ plus a relatively small number of rows representing fusion constraints. Because the scaling of this algorithm has a complicated depence on the constraint structure used, a general description of its runtime is difficult. However, in the case of multi-species network inference with one-to-one orthology, the network associated with each pair of orthologous genes requires solving a linear system with approximately twice as many observations and unknowns as the single species case. Linear systems of this size can be solved quickly using standard techniques. When the number of genes linked by fusion constraints becomes large, however, the augmented design matrix approach becomes impractical.

@@

\begin{equation}
\text{argmin}_{\beta^{S_i(t)}, \beta^{S_j(t)}} \displaystyle\sum_{S_i, S_j \in \text{species}} ||X^{S_i}\beta^{S_i}(t) - Y^{S_i}||^2 + \lambda_R||\beta^{S_i}(t)||^2 + \displaystyle \sum_{\beta_{g,k}^{S_i} \approx \beta_{h,l}^{S_j}} \lambda_S(\beta^{S_i}_{g,k}(t) - \beta_{h,l}^{S_j}(t-1)^2)
\end{equation}
Note that this is almost identical to equation 5, but now the network $\beta$ is a function of the iteration number $t$. On each step, the we compute $\beta$s that minimize a penalized cost function where the fusion penalty encourages similarity of a pair of parameters, using the previous iteration's solution. This process is iterated until the estimated $\beta$s converge. Because each iteration reduces the error between $\beta(t)$ and $\beta(t-1)$, and because $\beta(t) = \beta(t-1)$ is the globally optimal solution, this process must eventually converge to the same network as (cite 5). Although we have not produced bounds on the convergence rate, in practice a small number of iterations are necessary.

@@

With a saturating fusion penalty, fused interactions which appear to be very different based on expression data are allowed to unfuse from one another. A closely related problem has been studied in the context of LASSO regularization, where it was shown by Fan and Li that using a saturating penalty retains many of LASSO's desireable properties while removing its bias towards 0 \cite{fan2001variable}. They further showed that, although the resulting loss-function is nonconvex, good results can be obtained with a local quadratic approximation of gradient descent. Several saturating penalties, such as SCAD \cite{fan2001variable} and MCP \cite{zhang2010nearly}, have been discussed in the context of sparse regression. We introduce a modified form of MCP to the problem of penalizing differences between fused coefficients. The principle difference between the penalty we adopt and SCAD/MCP is that both of these penalties are L1 like at the origin, producing sparse solutions. Some network inference approaches use L1 penalties to produce sparse networks, on the basis that biological networks are thought to be sparse. However, since we are penalizing differences in interaction weights, rather than the weights themselves, there's no reason to assume that most differences will be exactly zero.

@@

$\beta^S(t+1)$ is obtained by fitting the ridge-fused model with fusion constraints given by the above $\lambda_S$. This is nice because all our penalties can be treated as L2 and therefore retain the properties of ridge regression.

Our adaptive penalty function introduces, in addition to regularization and fusion penalty weights $\lambda_R$ and $\lambda_S$, an unknown parameter $a$. We could employ grid search using cross-validation to searach for the best parameters, but for many data sets, this can be computationally expensive. Moreover, we are primarily interested in using this saturating penalty as a way of testing the hypothesis that conservation in GRNs can be predicted based off of known similarities between genes. Therefore, we implement a way of selecting $a$ to be equal to a certain percentile of the distribution of differences in fused interaction weights from the networks fit without fusion. This choice represents the user's hypothesis for the fraction of both networks that have unrelated function, and should therefore be unfused.

@@

\subsection{Simulated data}
We generate simulated data to evaluate the ability of our fused L2 approach to learn the true network and to show that sharing information between similar but not identical data sources results in more accurate network recovery. Because the purpose of adaptive fusion is to learn the structure of conservation while also learning networks, using simulated data allows us to test the method relative to a known gold standard for network conservation. 
Our simulations are based on the model $Y=X\beta+\sigma \epsilon$ where $\epsilion \sim N(0,1)$. Each data set consists of a training set and test set.

@@

Given a network $B$, we generate $N$ samples of gene expressions at each of two timepoints. The condition by gene expression matrix for timepoint one, $Y_{T1}$, is sampled randomly from a multivariate Gaussian distribution with identity covariance matrix. $X_{T1}$ is the TF expression sub-matrix of $Y_{T1}$, and consists of columns of $Y_{T1}$ that correspond to TFs. Treating the decay rate as 0, the gene expression matrix at timepoint two, $Y_{T2}$ is sampled as $Y_{T2} = Y_{T1} + BX_{T1}$. This process is carried out separately for each network.

Following generation of simulated data, we may introduce error into the orthology mapping. This can take the form of discarding a specified fraction of true orthologies (governed by a false-negative rate), by introducing random false orthologies (governed by a false-positive rate), or by adding Gaussian noise so that fused interactions are not identical. For convenience, the false-positive rate is specified in units of the number of true orthologs, and not the number of possible orthologs.

@@

\subsection{Beta scaling}
In previous work, betas were rescaled as to form a matrix of confidence scores $S$ as follows
\begin{equation}
S_{i,j} = \frac{\sigma^2_{\text{full model for }y_j}}{\sigma^2_{\text{full model for }y_j \text{ without predictor }i}}
\end{equation}
Computing residuals with respect to the data alone would disregard information gained through fusion. Instead, we used an approximation
\begin{equation}
S_{i,j} = \frac{\sigma^2_{\text{full model for }y_j}}{\sigma^2_{\text{full model for }y_j} + B_{i,j}^2 \times var(TF_j)}
\end{equation}
When the residual is uncorrelated with the prediction, then removing regressor $i$ with weight $B_{i,j}^2$ from the model of $j$ will increase the residual by $B_{i,j}^2$ times the variance of regressor $i$. In the case of regularized regression, the regressor and residual need not be uncorrelated, so the approximation will not hold exactly. This scaling is similar to rescaling according to variance explained relative to an augmented design matrix that includes fusion constraints. The intuition for scaling that a large $B_{i,j}$ may be inferred because of overfitting with a TF $i$ varies little across the available data, or because the regulatory weight is strong. Only in the latter case is a large weight suggestive of a true regulatory interaction.

>>>>

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become attainable \cite{bonneau_predictive_2007, ciofani_validated_2012, carro_transcriptional_2010}. Large-scale data collection efforts have contributed to the development of high quality networks which accurately recapitulate biological processes, but most processes and organisms remain uncharacterized at the network level. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine data from multiple platforms, lest we lose valuable information from existing studies. The problem of inferring related -- but not necessarily identical -- structure from related -- but not identical -- data is ubiquitous in biology. Multi-source network inference has applications for learning multiple networks in multiple related species, for learning networks associated with distinct processes within the same species, and for learning networks based on heterogenous data sources.

Moreover, as it becomes possible to learn genome-wide regulatory networks, we can begin to compare and to test whether there is conservation of networks across species and biological processes. Our use of model organisms to study biological processes and diseases relevant to humans relies on the assumption of conservation; yet this has not been effectively tested at the genome scale.

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference \cite{bonneau_predictive_2007, arrieta-ortiz_experimentally_2015}. The core of both methods is the observation that biological information about the relatedness of genes can be used to constrain which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Taking into account the similarity of putatively conserved interactions improves our ability to accurately describe TF-gene relationships on a genome-wide scale.

Our first method -- fused ridge -- uses an L2 penalty on the differences between a priori similar interactions (termed fusion penalty), and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when the overlap between genes is incomplete. Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously infer the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different). With this approach, we seek to test the hypothesis of conservation of interactions across networks.

@@

Existing multi species approaches often use orthology as a proxy for functional conservation \cite{penfold_inferring_2015, joshi_multi-species_2015, kashima_simultaneous_2009, zhang2010nearly}, or attempt to learn functional similarity via expression data \cite{gholami_cross-species_2010}. Orthology -- the measure of gene similarity we use -- can be approximated using readily identifiable sequence similarity, which is often a useful predictor of functional similarity \cite{wilson_assessing_2000}. In multi-species network inference, our fused L2 approach minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with evolutionary constraints created using orthology. However, we recognize that some genes may have taken on different functions and therefore may have new regulatory interactions. For example, gene duplications may lead to neofunctionalization \cite{eisen_phylogenomics:_1998} of the duplicated genes. Even in the case of inferring networks from multiple cell-lines, changes in chromatin configuration may affect our assumptions about the similarity of interactions \cite{li_role_2007}.

@@

We approach this problem by introducing a saturating penalty function based on work in statistics to develop unbiased regularization penalties for regression \cite{zhang2010nearly, fan2001variable}. The main difference between the L2 fusion approach and this adaptive fusion approach occurs when the difference between presumed analogous interactions is large despite the fusion penalty: we assume that large differences represent evidence that gene functions have diverged and represent this, in adaptive fusion, with a relaxation of the fusion penalty. The resulting cost function is non-convex and difficult to optimize \cite{fan2001variable}; however, we can approximate its solution and obtain deeper insight into functional similarity than is available through orthology or the comparison of separately learned networks.

Although the fusion constraints we employ can be described as arising from orthology - which links genes - it is important to note that the constraints themselves link individual regulatory interactions. This finer level of representational granularity is critical to the functioning of adaptive fusion, and means the method can accomodate any form of prior on expected similarity between regulatory interactions, even priors that are not decomposible into a gene to gene mapping. In order to solve the resulting optimization problem, we develop two algorithms for solving efficiently multi-output least-squares regression problems with pairwise L2 fusion penalties on entries of the coefficient matrix. We also introduce - in the form of adaptive fusion - the idea of a saturating penalty function on fusion constraints, and estimate the solution to the resulting optimization problem through iterative application of the fused L2 algorithm.

@@

Information about the conserved structure of gene regulation is introduced through the incorporation of constraints into the above regression formulation. These constraints penalize differences between interaction weights in the networks of multiple species that are expected to be similar based on prior biological knowledge. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $\lambda_Sp(\beta^A_{g,k} - \beta^B_{h,l})$ in the quantity being minimized in order to encourage similarity. The function $p(x)$ controls the shape of the relationship between weight dissimilarity and penalty, while parameter $\lambda_S$ controls the overall scaling of the penalization of differences between fused coefficients. $\lambda_S$ controls the tradeoff between fitting the expression data alone and producing a set of networks that conform to evolutionary prior knowledge. This gives us the final equation to be minimized

\begin{equation}
%\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} p_{\lambda_S}(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})
\argmin_{\beta^S} \displaystyle\sum_{S \in \{1, 2\}} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\substack{(g,h) \in orth,\\
 (h,l) \in orth}}p_{\lambda_S}(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})
\end{equation}

where the second sum is over pairs of interactions with fusion constraints. In the fused L2 algorithm we develop, the penalty function is equal to the L2 norm of the difference in regulatory weight of fused coefficients. That is, $p(x)=x^2$. The overall objective function for network inference is then:

@@

Because every component of the objective function is an L2 norm, the entire problem is convex, and can in fact be solved through linear regression with an augmented design matrix. %The regulators of a pair of genes can be solved independently as long as the pair is not linked by fusion constraints, or a chain of fusion constraints; thus the problem of cross-species GRN inference decomposes into a large number of manageable regression problems as long as the orthology mappings are sparse.

In order to better understand the behavior of this objective function, consider the case where there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism). The choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). In addition to allowing interpolation between these extremes, our method allows pooling of data even when there is incomplete orthology. By introducing constraints on the similarity of individual interactions, rather than on the networks as a whole \cite{liu2011temporal}, we can pool some information across species even when a small fraction of genes have orthologs. This is particularly useful when dealing with a large number of species; pairwise orthologies may be nearly complete even when the number of genes present in every organism is small.

\subsection{Statistical approach and background}
Most network inference problems have more predictors than experimental samples ($p >> N$). Many methods have been proposed for penalized regression, such as ridge regression \cite{hoerl_ridge_1970}, and applied to genomic data \cite{waldron_optimized_2011, li_network-constrained_2008}. In our application, we consider the case where related outputs (genes) are likely to be affected by common inputs (transcription factors). Fusion has been proposed as a method for utilizing correlation information from data by penalizing pairwise differences of coefficients \cite{Daye2009,Price2014}. Existing approaches using fusion have combined statistical strength across the multiple regression tasks by introducing regularization functions reflecting prior knowledge on output structure representing how outputs are related \cite{kim_tree-guided_2012,Land1997,Chen2010,Petry2011,Hebiri2011}. Price et al. and Bilgrau et al. use a fused ridge estimator for jointly estimating multiple inverse covariance matrices \cite{Price2014a,Bilgrau2015}; we take a similar approach, adding an fused L2 penalty to our ridge regression to incorporate prior knowledge about relationships between input-output pairs. In many cases, however, there is some uncertainty about the relationships that should be enforced. Sohn et al. attempt to simultaneously learn the regression coefficients and the output structure \cite{sohn_joint_2012}. We also take a simultaneous approach, where we apply a penalty function bounded by a constant to produce unbiased estimators for large coefficients, combined with an L2 penalty, similar to SCAD-L2 \cite{Zeng2012}. This approach allows us to simultaneously learn the regression coefficients and evaluate the validity of our prior information.

@@

We begin with the observation that ridge constraints can be incorporated into a least-squares regression problem by appending a scaled identity matrix to the design matrix, and a corresponding number of zeros to the response vector. Similarly, a fusion constraint $\lambda_S (\beta_{i} - \beta{j})^2$ can be incorporated into a least-squares regression problem by appending a row containing $\sqrt{\lambda_S}$ in the $i$th position, $-\sqrt{\lambda_S}$ in the $j$th position, and $0$s elsewhere to the design matrix, and zero to the response vector.

In order to convert an optimization over multiple response variables (indexed by $i$) and multiple sources (indexed by $j$) with design matrices $X^{S_j}$ and response variables $y_i^{S_j}$ into an optimization with a single source and response variable, we vectorize as follows. Given a set containing $(i,j)$ for each of the data sources and response variables currently being optimized, we construct the matrix $A$ by diagonally concatenating each of $X^{S_j}$ (with the remaining entries 0) and vertically concatenating each of $y_j^{S_j}$. We then solve
%Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
%$y =$ vertically concatenate the columns of $Y$. We solve

@@

%We can then define $W_{i,j} = \beta_{(j-1) \times N + i, (j-1) \times M + j}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication. In an ordinary regression problem each response variable can be solved independently, and vectorization is unnecessary. However, in fused regression, we append additional rows to the design matrix $A$ that link entries of $W$ associated with different response variables. As a result, these linked response variables must be solved simultaneously through vectorization. Two response variables are linked by a fusion constraint if any of the regulatory weights affecting those genes are linked by a fusion constraint. Two response variables must be solved simultaneously if there is any chain of linked response variables connecting them. However, every other response variable can be solved separately. In biological terms, the regulators of two genes (whether in the same species, or different species) must be solved together if there is a fusion constraint linking those genes' regulators, or if there is a chain of such constraints. This is important because vectorization produces a design matrix with a number of rows and columns equal to the sum of the number of rows and columns of the design matrices associated with each of the response variables (genes) being solved for. If the networks for a large number of genes are solved simultaneously, the system of equations can quickly become intractible to solve.

In order to avoid this difficulty, we use depth-first search to identify linked columns of each TF expression matrix $Y^{S_j}$, then forming design and response matrices through vectorization. We can then incorporate fusion constraints as in the case of single-source single response-variable fused regression. In most cases, we have found the direct solution using augmented matrices to be adequate. This is because the structure of orthology tends to be sparse, so only a small number of genes must be solved at once. However, in the general case of $c_i$ response variables from each of $i \in \text{species}$ data sources the design matrix to be solved is of size $\displaystyle \sum_{c_i} rows(X^{S_i}) \times \displaystyle \sum_{c_i} cols(X^{S_i})$ plus a relatively small number of rows representing fusion constraints. Because the scaling of this algorithm has a complicated depence on the constraint structure used, a general description of its runtime is difficult. However, in the case of multi-species network inference with one-to-one orthology, the network associated with each pair of orthologous genes requires solving a linear system with approximately twice as many observations and unknowns as the single species case. Linear systems of this size can be solved quickly using standard techniques. When the size of the groups of genes linked by fusion constraints becomes large, however, the augmented design matrix approach becomes impractical.

@@

\begin{equation}
\text{argmin}_{\beta^{S_i(t)}, \beta^{S_j(t)}} \displaystyle\sum_{S_i, S_j \in \text{species}} ||X^{S_i}\beta^{S_i}(t) - Y^{S_i}||^2 + \lambda_R||\beta^{S_i}(t)||^2 + \displaystyle \sum_{\beta_{g,k}^{S_i} \approx \beta_{h,l}^{S_j}} \lambda_S(\beta^{S_i}_{g,k}(t) - \beta_{h,l}^{S_j}(t-1)^2)
\end{equation}
Note that this is almost identical to equation 5, but now the network $\beta$ is a function of the iteration number $t$. On each step, the we compute $\beta$s that minimize a penalized cost function where the fusion penalties encourage similarity between a parameter and its fused-to parameter from the previous iteration's solution. This process is iterated until the estimated $\beta$s converge. Because each iteration reduces the error between $\beta(t)$ and $\beta(t-1)$, and because $\beta(t) = \beta(t-1)$ is the globally optimal solution, this process must eventually converge to the same network as (cite 5). Although we have not produced bounds on the convergence rate, which also depends on the structure of constraints, in practice a small number of iterations are necessary.

@@

With a saturating fusion penalty, fused interactions which appear to be very different based on expression data are allowed to unfuse from one another. A closely related problem has been studied in the context of LASSO regularization, where it was shown by Fan and Li that using a saturating penalty retains many of LASSO's desireable properties while removing its bias towards 0 \cite{fan2001variable}. They further showed that, although the resulting loss-function is nonconvex, good results can be obtained with a local quadratic approximation of gradient descent. Several saturating penalties, such as SCAD \cite{fan2001variable} and MCP \cite{zhang2010nearly}, have been discussed in the context of sparse regression. We introduce a modified form of MCP to the problem of penalizing differences between fused coefficients. The principle difference between the penalty we adopt and SCAD/MCP is that both of these penalties are L1 like at the origin, producing sparse solutions. Some network inference approaches use L1 penalties to produce sparse networks, on the basis that biological networks are thought to be sparse. However, since we are penalizing differences in interaction weights, rather than the weights themselves, there's no reason to assume that most differences will be exactly zero, and an L2 penalty - equivalent to an assumption that the differences between fused coefficients are gaussian distributed - may be more appropriate.

@@

$\beta^S(t+1)$ is obtained by fitting the ridge-fused model with fusion constraints given by the above $\lambda_S$. This is useful because all our penalties can be treated as L2 and therefore retain the properties of ridge regression, and can be solved using the fused L2 algorithm we develop.

Our adaptive penalty function introduces, in addition to regularization and fusion penalty weights $\lambda_R$ and $\lambda_S$, an unknown parameter $a$. We could employ grid search using cross-validation to searach for the best parameters, but for many data sets, this can be computationally expensive. Moreover, we are primarily interested in using this saturating penalty as a way of testing the hypothesis that conservation in GRNs can be predicted based off of known similarities between genes. Therefore, we propose a method for selecting $a$ to be equal to a certain percentile of the distribution of differences in fused interaction weights from the networks fit without fusion. This choice represents the user's hypothesis for the fraction of both networks that have unrelated function, and should therefore be unfused.

@@

\subsection{Simulated data}
We generate simulated data to evaluate the ability of our fused L2 approach to learn the true network and to show that sharing information between similar but not identical data sources results in more accurate network recovery. Because the purpose of adaptive fusion is to learn the structure of conservation while also learning networks, using simulated data allows us to test the method relative to a known gold standard for network conservation.

@@

Given a network $B$, we generate $N$ samples of gene expressions at each of two timepoints. The condition by gene expression matrix for timepoint one, $Y_{T1}$, is sampled randomly from a multivariate Gaussian distribution with identity covariance matrix. $X_{T1}$ is the TF expression sub-matrix of $Y_{T1}$, and consists of columns of $Y_{T1}$ that correspond to TFs. Treating the decay rate as 0, the gene expression matrix at timepoint two, $Y_{T2}$ is sampled as $Y_{T2} = Y_{T1} + BX_{T1} + \epsilon$, where $\epsilon$ is a Gaussian noise term. This process is carried out separately for each network.

Following generation of simulated data, we may introduce error into the orthology mapping. This can take the form of discarding a specified fraction of true orthologies (governed by a false-negative rate), by introducing random false orthologies (governed by a false-positive rate), or by adding Gaussian noise so that fused interactions are not identical (described above). For convenience, the false-positive rate is specified in units of the number of true orthologs, and not the number of possible orthologs.

@@

\subsection{Beta scaling}
In previous work, betas were rescaled as to form a matrix of confidence scores $S$ as follows
\begin{equation}
S_{i,j} = \frac{\sigma^2_{\text{full model for }y_j}}{\sigma^2_{\text{full model for }y_j \text{ without predictor }i}}
\end{equation}
Computing residuals with respect to the data alone would disregard information gained through fusion, because certain iteractions may be large due to fusion, rather than their individual explanatory power. Instead, we used an approximation
\begin{equation}
S_{i,j} = \frac{\sigma^2_{\text{full model for }y_j}}{\sigma^2_{\text{full model for }y_j} + B_{i,j}^2 \times var(TF_j)}
\end{equation}
When the residual is uncorrelated with the prediction, then removing regressor $i$ with weight $B_{i,j}^2$ from the model of $j$ will increase the residual by $B_{i,j}^2$ times the variance of regressor $i$. In the case of regularized regression, the regressor and residual need not be uncorrelated, so the approximation will not hold exactly. This scaling is similar to rescaling according to variance explained relative to an augmented design matrix that includes fusion constraints. The intuition for scaling that a large $B_{i,j}$ may be inferred because of overfitting with a TF $i$ varies little across the available data, or because the regulatory weight is strong. Only in the latter case is a large weight suggestive of a true regulatory interaction.
